# Summary statistics and tidy data {#sec-normalization}

Last time we learned the basics of reading writing, and manipulating data. To review, download `pop_data.csv` (from [this link](https://github.com/dysordys/data-with-R/tree/main/data)) and set your working directory to the folder where you saved it. This is a comma-separated file (CSV), so you can load it using `read_csv`. As usual, we first load the `tidyverse` package:

```{r}
#| message: false
library(tidyverse)
```

And now we may use the `tidyverse` functionalities, such as `read_csv`:

```{r}
#| message: false
pop <- read_csv("pop_data.csv")
```

The data we just loaded contain population densities of three species at two spatial patches (A and B) at various points in time, ranging from 1 to 50 in steps of 1:

```{r}
pop
```

One can now perform various manipulations on these data, by using the functions `rename`, `select`, `filter`, `arrange`, and `mutate` we have learned about in @sec-wrangling. For instance, we could create a new column called `total` which contains the total community density (sum of the three species' population densities) at each point in time and each location:

```{r}
mutate(pop, total = species1 + species2 + species3)
```

As a reminder, this can be written equivalently, using pipes:

```{r}
pop %>% mutate(total = species1 + species2 + species3)
```


## Creating summary data

One can create summaries of data using the `summarise` function. This will simply apply some function to a column. For example, to calculate the average population density of species 1 in `pop`, across both time and patches, one can write

```{r}
pop %>% summarise(meanDensity1 = mean(species1))
```

Here `meanDensity1` is the name of the new column to be created, and the `mean` function is our summary function, collapsing the data into a single number.

So far, this is not particularly interesting; in fact, the exact same effect would have been achieved by typing the shorter `mean(pop$species1)` instead. The real power of `summarise` comes through when combined with `group_by`. This groups the data based on the given grouping variables. Let us see how this works in practice:

```{r}
#| message: false
pop %>% group_by(patch)
```

Seemingly nothing has happened; the only difference is the extra line of comment above, before the printed table, saying `Groups: patch [2]`. What this means is that the rows of the data were internally split into two groups. The first have `"A"` as their patch, and the second have `"B"`. Whenever one groups data using `group_by`, rows which share the same unique combination of the grouping variables now belong together, and *subsequent* operations will act separately on each group instead of acting on the table as a whole (which is what we have been doing so far). That is, `group_by` does not actually alter the data; it only alters the behaviour of the functions applied to the grouped data.

If we group not just by `patch` but also by `time`, the comment above the table will read `Groups: patch, time [100]`:

```{r}
#| message: false
pop %>% group_by(patch, time)
```

This is because there are 100 unique combinations of patch and time: two different `patch` values (`"A"` and `"B"`), and fifty points in time (1, 2, ..., 50). So we have "patch A, time 1" as group 1, "patch B, time 1" as group 2, "patch A, time 3" as group 3, and so on until "patch B, time 50" as our group 100.

As mentioned, functions that are applied to grouped data will act on the groups separately. To return to the example of calculating the mean population density of species 1 in the two patches, we can write:

```{r}
#| message: false
pop %>%
  group_by(patch) %>%
  summarise(meanDensity1 = mean(species1))
```

One may obtain multiple summary statistics within the same `summarize` function. Below we compute both the mean and the standard deviation of the densities per patch:

```{r}
#| message: false
pop %>%
  group_by(patch) %>%
  summarise(meanDensity1 = mean(species1), sdDensity1 = sd(species1))
```

Let us see what happens if we calculate the mean density of species 1---but grouping by `time` instead of `patch`:

```{r}
#| message: false
pop %>%
  group_by(time) %>%
  summarise(meanDensity1 = mean(species1))
```

The resulting table has 50 rows---half the number of rows in the original data, but many more than the two rows we get after grouping by `patch`. The reason is that there are 50 unique time points, and so the average is now computed over those rows which share `time`. But there are only two rows per moment of time: the rows corresponding to patch A and patch B. When we call `summarise` after having grouped by `time`, the averages are computed over the densities in these two rows only, per group. That is why here we end up with a table which has a single row per point in time.


:::{.callout-warning}
One common mistake when one first meets with grouping and summaries is to assume that if we call `group_by(patch)`, then the subsequent summaries will be taken over patches. *This is not the case*, and it is important to take a moment to understand why. When we apply `group_by(patch)`, we are telling R to treat different patch values as group indicators. Therefore, when creating a summary, only the patch identities are retained from the original data (apart from the new summary statistics we calculate, of course). This means that the subsequent summaries are taken over everything *except* the patches. This should be clear after comparing the outputs of

```{r}
#| eval: false
pop %>% group_by(patch) %>% summarise(meanDensity1 = mean(species1))
```

and

```{r}
#| eval: false
pop %>% group_by(time) %>% summarise(meanDensity1 = mean(species1))
```

The first distinguishes the rows of the data only by `patch`, and therefore the average is taken over time. The second distinguishes the rows by `time`, so the average is taken over the patches. Run the two expressions again to see the difference between them!
:::


We can use functions such as `mutate` or `filter` on grouped data. For example, we might want to know the deviation of species 1's density from its average *in each patch*. Doing the following does not quite do what we want:

```{r}
pop %>% mutate(species1Dev = species1 - mean(species1))
```

This will put the difference of species 1's density from its mean density across both time and patches into the new column `species1Dev`. That is not the same as calculating the difference from the mean in a given patch---patch A for rows corresponding to patch A, and patch B for the others. To achieve this, all one needs to do is to group the data by `patch` before invoking `mutate`:

```{r}
pop %>%
  group_by(patch) %>%
  mutate(species1Dev = species1 - mean(species1))
```

Comparing this with the previous table, we see that the values in the `species1Dev` column are now different, because this time the differences are taken with respect to the average densities per each patch.

Finally, since `group_by` changes subsequent behaviour, we might eventually want to get rid of the grouping in our data. This can be done with the function `ungroup`. For example:

```{r}
#| message: false
pop %>%
  group_by(patch) %>%
  summarise(meanDensity1 = mean(species1), sdDensity1 = sd(species1)) %>%
  ungroup()
```

It is good practice to always `ungroup` the data after we have calculated what we wanted using the group structure. Otherwise, subsequent calculations could be influenced by the grouping in unexpected ways.


## Tidy data {#sec-tidy}

In science, we often strive to work with *tidy data* (also known as *long-format data* or *normalized data*). A dataset is tidy if:

1. Each variable is in its own column;
2. Each observation is in its own row.

Tidy data are suitable for performing operations, statistics, and plotting on. Furthermore, tidy data have a certain well-groomed feel to them, in the sense that their organization always follows the same general pattern regardless of the type of dataset one studies. (By contrast, every non-tidy dataset tends to be messy in its own unique way.) The `tidyverse` offers a simple and convenient way of putting data in tidy format.

For example, the `pop` table from the previous section is not tidy. This is because although each variable is in its own column, it is not true that each observation is in its own row. In fact, each row contains three observations: the densities of species 1, 2, and 3 at a given time and place. To tidy up these data, we create *key-value pairs*. We merge the columns for species densities into just two new ones. The first of these (the *key*) indicates whether it is species 1, or 2, or 3 which the given row refers to. The second column (the *value*) contains the population density of the given species. Such key-value pairs are created by the function `pivot_longer`:

```{r}
pop %>% pivot_longer(cols = 3:5, names_to = "species", values_to = "density")
```

The function `pivot_longer` takes three arguments in addition to the first (data) argument that we may pipe in, like above. First, `cols` is the list of columns to be converted into key-value pairs. One can refer to the columns by number: `3:5` is the same as `c(3, 4, 5)` and selects the third, fourth, and fifth columns---the ones corresponding to the population densities. We could also have written `c("species1", "species2", "species3")` instead, choosing columns by their names. This can give greater clarity, albeit at the cost of more typing. Second, the argument `names_to` is the name of the new key column. Finally, `values_to` is the name of the new value column.

Notice that the above table is now in tidy form: each column records a single variable, and each row contains a single observation. Notice also that, unlike the original `pop` which had 100 rows and 5 columns, the tidy version has 300 rows and 4 columns. This is natural: since the number of rows was reduced, there must be some extra rows to prevent the loss of information.

It is possible to "undo" the effect `pivot_longer`. To do so, use `pivot_wider`:

```{r}
pop %>%
  pivot_longer(cols = 3:5, names_to = "species", values_to = "density") %>%
  pivot_wider(names_from = "species", values_from = "density")
```

The two named arguments of `pivot_wider` above are `names_from` (which specifies the column from which the names for the new columns will be taken), and `values_from` (the column whose values will be used to fill in the rows under those new columns).

As a remark, one could make the data even "wider", by not only making columns out of the population densities, but the densities at a given patch. Doing so is simple: one just needs to specify both the `species` and `patch` columns from which the new column names will be compiled.

```{r}
pop %>%
  pivot_longer(cols = 3:5, names_to = "species", values_to = "density") %>%
  pivot_wider(names_from = c("species", "patch"), values_from = "density")
```

If tidy data are what we strive for, what is the practical use of `pivot_wider`? There are two answers to this question. First, while non-tidy data are indeed less efficient from a computational and data analysis standpoint, they are often more human-readable. For example, the `pop` table is easy to read despite the lack of tidiness, because each row corresponds to a given time and place. By tidying the data, information referring to any given time and place will be spread out over multiple (in our case, three) rows---one for each species. While this is preferable from a data analysis point of view, it can be more difficult to digest visually. Second, wide data lend themselves very well to a class of statistical techniques called *multivariate analysis*. In case one wants to perform multivariate analysis, wide-format data are often better than tidy data. We will see an example application of this in @sec-ancova.

Finally, it is worth noting the power of tidy data in, e.g., generating summary statistics. To obtain the mean and the standard deviation of the population densities for each species in each patch, all one has to do is this:

```{r}
#| message: false
pop %>%
  # Tidy up the data:
  pivot_longer(3:5, names_to = "species", values_to = "density") %>%
  # Group data by both species and patch:
  group_by(patch, species) %>%
  # Obtain statistics:
  summarise(meanDensity = mean(density), sdDensity = sd(density)) %>%
  # Don't forget to ungroup the data at the end:
  ungroup()
```


## Exercises

The exercises below use the `iris` dataset---the same that we used in the previous chapter's data wrangling exercises (@sec-wrangling-exercises). Convert the `iris` data to a tibble with the `as_tibble()` function, and assign it to a variable.

1. Create a new column in the `iris` dataset which contains the deviation of petal lengths from the average of the whole dataset.

2. Create a new column in the `iris` dataset which contains the deviation of petal lengths from the average *of each species*. (Hint: `group_by` the species and then `mutate`!)

3. Create a table where the rows are the three species, and the columns are: average petal length, variance of petal length, average sepal length, and variance of sepal length.

4. Create key-value pairs in the `iris` dataset for the petal characteristics. In other words, have a column called `Petal.Trait` (whose values are either `Petal.Length` or `Petal.Width`), and another column called `Petal.Value` (with the length/width values).

5. Repeat the same exercise, but now for sepal traits.

6. Finally, do it for both petal and sepal traits simultaneously, to obtain a fully tidy form of the `iris` data. That is, the key column (call it `Flower.Trait`) will have the values `Petal.Length`, `Petal.Width`, `Sepal.Length`, and `Sepal.Width`. And the value column (which you can call `Trait.Value`) will have the corresponding measurements.
